Fail Faster

### 2025-03-17 paper/paper.tex
paper blah

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{Why is a monadic generator library slow?}
\jwc{NOTE: We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}

@@

\subsection{Monadic DSL Abstraction Overhead}

\jwc{
  \begin{itemize}
    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = (Bq.gen_int 0 100) in
  let%bind y = (Bq.gen_int 0 x) in
  Bq.return (x,y)

let int_pair_inlined : (int * int) Bq.t
  fun sr ->
    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
    let y = Splittable_random.int sr ~lo:0 ~hi:x in
    (x,y)
\end{lstlisting}

\jwc{Show benchmark difference between these two generators: the benchmark is in \texttt{waffle-house/handwritten-ocaml/bin/basic-compare.ml}. It's about 2x. (see comment in latex here.)}

@@

\jwc{The overhead of the abstraction is 2x.
The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}

\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}

\begin{lstlisting}
def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\end{lstlisting}

\jwc{Same story in scala. 
Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
}

>>>>

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{In this section we answer: ``Why are monadic generator libraries slow?''}

@@

\subsection{Monadic DSL Abstraction Overhead}
Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
\jwc{Should we do this?}

Different PBT libraries use variations on this basic design. Haskell's QuickCheck uses a reader monad with an immutable state
that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
\jwc{However, the principles are the same, and we will use \bq{} to illustrate.}

\begin{figure}
\begin{lstlisting}
module Bq = struct
  type 'a t = int -> SR.t -> 'a

  let return (x : 'a) : 'a t = fun _ _ -> x

  let bind (g : 'a t) (k : 'a -> 'b t) : 'b t =
    fun size random ->
      let a = g size random in
      (k a) size random

  let gen_int (lo : int) (hi : int) : int t =
    fun _ random -> SR.int random lo hi
end
\end{lstlisting}
\caption{Internals of a Monadic Generator eDSL}
\label{fig:bq-internals}
\end{figure}

Just how much runtime overhead does this monadic abstraction introduce? To
illustrate, let's return to our running example of a constrained pair of
integers, written in both \bq{} and ScalaCheck.
Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
The first version (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) is written with monadic generator combinators.
\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
The second version (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
is just the first version, with (a) all generator combinator definitions inlined, and (b)
beta-redexes reduced.

\begin{figure}
\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = (Bq.gen_int 0 100) in
  let%bind y = (Bq.gen_int 0 x) in
  Bq.return (x,y)

let int_pair_inlined : int -> SR.t -> int * int =
  fun _ sr ->
    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
    let y = Splittable_random.int sr ~lo:0 ~hi:x in
    (x,y)

def intPair : Gen[(Long,Long)] = for {
  x <- Gen.choose(0,1000)
  y <- Gen.choose(0,x)
} yield (x,y)

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\caption{Int Pair Generators in \bq{} and ScalaCheck}
\label{fig:overhead-explanation-code}
\end{figure}

The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
languages, the inlined version takes \textbf{half} as much time to generate a
single pair of \texttt{int}s.

\begin{figure}
\end{lstlisting}
  \begin{tabular}{lll}
  Library    & Generator          & Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\label{fig:overhead-explanation-perf}
\end{figure}

@@

% \jwc{The overhead of the abstraction is 2x. The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}

% \jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}

% \jwc{Same story in scala. 
% Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
% This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
% }
